SDSS-IV MaNGA: The Radial Profile of Enhanced Star Formation in Close Galaxy Pairs

### 2020-10-01 maintex.tex
minor updates

\begin{abstract}
We compare the radial profiles of the specific star formation rate (sSFR) in a sample of 220 star-forming galaxies in close pairs with those of mass-matched isolated galaxies in the SDSS-IV MaNGA survey. We find that the sSFR is centrally enhanced (within one effective radius) in interacting galaxies by $\sim$0.3 dex and that there is a weak sSFR suppression in the outskirts of the galaxies of $\sim$0.1 dex. We stack the differences profiles for galaxies in five stellar mass bins between \logm\ $=$ 9.0$-$11.5 and find that the sSFR enhancement has no dependence on the stellar mass. The same result is obtained when the comparison galaxies are matched to each paired galaxy in both stellar mass and redshift. In addition, we find that that the sSFR enhancement is elevated in pairs with nearly equal masses and closer projected separations, in agreement with previous work based on single-fiber spectroscopy.
\end{abstract}

@@

We classify galaxies in this survey as star forming galaxies using the color-magnitude diagram (CMD). We show the CMD for the MaNGA survey and our pair sample in Figure \ref{fig:cmd} along with demarcation lines which separate the blue cloud, red sequence, and green valley. We established the demarcation lines by collapsing the CMD to a color histogram for each of the three regions. We then varied the slopes between the regions until we found the slopes which best fit the data. These demarcation lines are;

@@

Where $NUV-r$ is the color from SDSS's k-corrected absolute magnitude and $\mathcal{M}_i$ is the i-band magnitude from the NSA catalog. We have checked the selected star forming galaxies on the BPT diagram \citep{Baldwin:1981} to confirm that AGN (active galactic nuclei) will have a minimal impact on our SFG sample.

On top of the color selection cuts, we require that all galaxies are in the Primary or Secondary MaNGA subsamples \citep{Wake:2017}, and that the stellar mass range of the galaxy sample is between \logm\ $=$ 9.0$-$11.5.

@@

Several of the target galaxies in our sample have multiple companions. From the whole set of 404 IFUs, there are 327 pairs, 67 triplets, 7 quadruplets, and 1 quintuplet.

@@

From the NSA catalog's 641,409 galaxies, we find 492 galaxies which are paired to MaNGA targets. MaNGA targets which have both an inside-IFU and an outside-IFU pair are left to the inside-IFU sample. After restricting the sample to SFGs, we have another 133 MaNGA targets with paired galaxies outside of the IFU. This give use a total of 220 MaNGA targets in our pair sample.

@@

To show how the population of galaxy pairs differs from isolated galaxies, we will compare our pair sample to a sample of isolated galaxies in the MaNGA survey. We build this control sample from the MaNGA target galaxies which have no spectroscopic companions within $r_p$ $<$ 50 kpc and $\Delta v$ $<$ 500 km s$^{-1}$ either inside or outside of the IFU. This gives us a control sample of 1891 star forming control galaxies.

@@

Setting the above requirements, most paired galaxies will find between 20$-$100 control galaxies. Since we want each paired galaxy to be treated in a similar manner, we will down-select the total number of acquired control galaxies down to 20 controls randomly. In the cases where a paired galaxy does not initially acquire 20 control galaxies we iteratively expand the stellar mass limit by 0.1 dex until at least 20 control galaxies are found. A small set of paired galaxies required extra iterations to find 20 controls; 8 needed an extra iteration, 1 needed two extra iterations, and 1 needed three iterations.

@@

It has been shown that hot low mass evolved stars (HOLMES) in retired galaxies can mimic H\,{\sc ii} regions and AGN on the BPT diagram \citep{Stasinska:2008}. To mask retired spaxels, we implement a H$\alpha$ equivalent width (EW) cut of \ewha\ $>$ 3\AA\ \citep{Cid-Fernandes:2011}.

@@

We fit the geometry with apertures from both catalog when available. When using the first method of pair-control comparison, where pairs and controls are grouped into stellar mass bins, we fit the control galaxies with the NSA apertures. When making the comparison with the second method, paired galaxies fitted with the NSA aperture are compared against controls fitted with the NSA aperture and paired galaxies fitted with the \citet{Simard:2011} apertures are compared against controls fitted with the \citet{Simard:2011} apertures.

We compared the magnitudes extracted from the two apertures for the control galaxies and found that they are in close agreement. This means that the geometries between the two surveys must be in close agreement and the profiles of a galaxy fitted with a NSA aperture can be confidently compared against a galaxy fitted with an aperture from \citet{Simard:2011}.

@@

Between the median profiles of the control galaxies and paired galaxies, we can see that the paired profiles are flattened in their centers in comparison to the control profiles. The paired profiles are centrally enhanced by $\sim$0.25 $\pm$ 0.1 dex. This enhancement linearly falls to zero around 1.0 $R$/\reff. In the disks the profiles feature a weak suppression to the sSFR of 0.05$-$0.1 dex.

@@

Once the average profile of the control galaxies is made, we take the difference between the paired galaxy's profile and the average control profile in log space. We do this process for each of the 220 paired galaxies.

@@

In this work, we studied a sample of paired, star forming galaxies using the MaNGA IFS survey. We have made a sample of 220 star forming galaxy pairs along with a set of 1891 star forming control galaxies. We compared the radial profiles of the paired galaxies with those of the controls galaxies and found the follwing;

\begin{enumerate}
\item The sSFR is centrally enhanced within the inner 1.0 R/\reff\ by 0.25 $\pm$ 0.1 dex. 
\item The disks of the paired galaxies, R/\reff\ $=$ 1.0$-$2.5, show limited suppression in their disks, $\sim$0.1 dex.
\item The sSFR enhancement in paired galaxies is independent of the total stellar mass of the paired galaxies, Massive galaxies experience the same level of sSFR enhancement as low mass galaxies.
\item The sSFR enhancement in paired galaxies does depend on the ratio between the masses of the paired galaxies. Galaxies with small mass ratios, $|\Delta$log(M)$|$ $=$ 0.0$-$0.5, see an enhancement to sSFR of 0.3 $\pm$ 0.1 dex, which is 0.1 dex higher than the average paired galaxies in the survey. We have also showed that paired galaxies with large mass ratios, $|\Delta$log(M)$|$ $=$ 1.0, still show substantial levels of enhancement to the sSFR, 0.15$-$0.2 dex. 
\item Further, in paired galaxies with large mass ratios, the less massive galaxies experience higher levels of sSFR enhancement than the more massive galaxy in the pair, $\sim$0.1 dex.
\item The sSFR enhancement increases with closer projected separations where galaxies within 5$-$15 kpc have their sSFR enhanced by 0.3$-$0.4 $\pm$ 0.1 dex. 
\item When splitting the sSFR enhancement profiles by stellar mass, mass ratio, or projected separation, the general shape of the enhancement profile is unchanged. While lower projected separations or mass ratios closer to 1:1 may raise the level of the central enhancement to the profiles, the profiles still linearly fall to zero enhancement around 1.0 R/\reff\ with roughly the same gradient.
\item The paired galaxies will experience a fractional mass growth rate of 1$-$6\%, where the growth in mass in largest in low mass galaxies.

>>>>

\begin{abstract}
We compare the radial profiles of the specific star formation rate (sSFR) in a sample of 175 star-forming galaxies in close pairs with those of mass-matched isolated galaxies in the SDSS-IV MaNGA survey. We find that the sSFR is centrally enhanced (within one effective radius) in interacting galaxies by $\sim$0.3 dex and that there is a weak sSFR suppression in the outskirts of the galaxies of $\sim$0.1 dex. We stack the differences profiles for galaxies in five stellar mass bins between \logm\ $=$ 9.0$-$11.5 and find that the sSFR enhancement has no dependence on the stellar mass. The same result is obtained when the comparison galaxies are matched to each paired galaxy in both stellar mass and redshift. In addition, we find that that the sSFR enhancement is elevated in pairs with nearly equal masses and closer projected separations, in agreement with previous work based on single-fiber spectroscopy.
\end{abstract}

@@

We classify galaxies in this survey as star forming galaxies by selecting galaxies in the blue cloud on the color-magnitude diagram (CMD). We show the CMD for the MaNGA survey and our pair sample in Figure \ref{fig:cmd} along with demarcation lines which separate the blue cloud, red sequence, and green valley. We established the demarcation lines by collapsing the CMD to a color histogram for each of the three regions. We then varied the slopes between the regions until we found the slopes which best fit the data. These demarcation lines are;

@@

Where $NUV-r$ is the color from SDSS's k-corrected absolute magnitude and $\mathcal{M}_i$ is the i-band magnitude from the NSA catalog. We further apply an \ewha\ $\ge$ 3\AA\ cut on galaxy spectra extracted from a 1.0 \reff\ aperture to remove any passive galaxies which may have passed the CMD cut \citep{Cid-Fernandes:2011}. We further use the BPT diagnostic \citep{Baldwin:1981} to remove galaxies with an AGN from our sample. We use the emission line ratios, log([O\textsc{III}]/H$\beta$) and log([N\textsc{II}]/H$\alpha$), extracted from a 1.3 kpc aperture along with the maximum starburst line from \citet{Kewley:2001} as the demarcation between star forming branch and the AGN branch.

On top of the star formation cuts, we require that all galaxies are in the Primary or Secondary MaNGA subsamples \citep{Wake:2017}, and that the stellar mass range of the galaxy sample is between \logm\ $=$ 9.0$-$11.5. We use masses calculated from the elliptical Petrosian apertures in the NSA catalog for our total stellar masses.

@@

Several of the target galaxies in our sample have multiple companions. From the whole set of 404 IFUs, there are 327 pairs, 67 triplets, 7 quadruplets, and 1 quintuplet. In cases where a MaNGA target has multiple viable companions galaxies, we chose the closest companion when measuring the mass ratio and projected separation.

@@

From the NSA catalog's 641,409 galaxies, we find 492 galaxies which are paired to MaNGA targets. MaNGA targets which have both an inside-IFU and an outside-IFU pair are left to the inside-IFU sample. After restricting the sample to SFGs, we have another 133 MaNGA targets with paired galaxies outside of the IFU. This give use a total of 175 MaNGA targets in our pair sample.

@@

To show how the population of galaxy pairs differs from isolated galaxies, we will compare our pair sample to a sample of isolated galaxies in the MaNGA survey. We build this control sample from the MaNGA target galaxies which have no spectroscopic companions within $r_p$ $<$ 50 kpc and $\Delta v$ $<$ 500 km s$^{-1}$ either inside or outside of the IFU. This gives us a control sample of 1888 star forming control galaxies.

@@

Setting the above requirements, most paired galaxies will find between 20$-$100 control galaxies. Since we want each paired galaxy to be treated in a similar manner, we will down-select the total number of acquired control galaxies down to 20 controls randomly. Control galaxies may be selected for multiple pairs by the pipeline; most of the control galaxies are used at least once and the average number of times a given control is reused is between 4$-$6 times. In the cases where a paired galaxy does not initially acquire 20 control galaxies we iteratively expand the stellar mass limit by 0.1 dex until at least 20 control galaxies are found. Only 5 of the paired galaxies required an extra iteration to find 20 controls.

@@

@@

We fit the geometry with apertures from both catalogs when available. When using the first method of pair-control comparison, where pairs and controls are grouped into stellar mass bins, we fit the control galaxies with the NSA apertures. When making the comparison with the second method, paired galaxies fitted with the NSA aperture are compared against controls fitted with the NSA aperture and paired galaxies fitted with the \citet{Simard:2011} apertures are compared against controls fitted with the \citet{Simard:2011} apertures. Further, when defining the a galaxy's geometry with the \citet{Simard:2011} apertures we also use the masses given in the catalog for the total stellar mass of the galaxy.

@@

The profiles of the control galaxies are centrally quenched, which was previously seen in the MaNGA sample in \citet{Belfiore:2018}. The profiles of the paired galaxies flat across the whole galaxy accept for galaxies with masses above \logm\ $=$ 10.5 which show some central quenching. When taking the difference between the pairs and controls, the paired profiles are centrally enhanced by $\sim$0.25 $\pm$ 0.1 dex. This enhancement linearly falls to zero around 1.0 $R$/\reff. In the disks the profiles feature a weak suppression to the sSFR of 0.05$-$0.1 dex.

@@

Once the average profile of the control galaxies is made, we take the difference between the paired galaxy's profile and the average control profile in log space. We do this process for each of the 175 paired galaxies.

@@

In this work, we studied a sample of paired, star forming galaxies using the MaNGA IFS survey. We have made a sample of 175 star forming galaxy pairs along with a set of 1888 star forming control galaxies. We compared the radial profiles of the paired galaxies with those of the controls galaxies and found the following;

\begin{enumerate}
\item The sSFR is centrally enhanced within the inner 1.0 R/\reff\ by 0.25 $\pm$ 0.1 dex. 
\item The outskirts of the paired galaxies, R/\reff\ $=$ 1.0$-$2.5, show limited suppression in their disks of $\sim$0.1 dex.
\item The sSFR enhancement in paired galaxies is independent of the total stellar mass of the paired galaxies, Massive galaxies experience the same level of sSFR enhancement as low mass galaxies.
\item The sSFR enhancement in paired galaxies does depend on the ratio between the masses of the paired galaxies. Galaxies with small mass ratios, $|\Delta$log(M)$|$ $=$ 0.0$-$0.5, see an enhancement to sSFR of 0.3 $\pm$ 0.1 dex, which is 0.1 dex higher than the average paired galaxies in the survey. We have also showed that paired galaxies with large mass ratios, $|\Delta$log(M)$|$ $=$ 1.0, still show substantial levels of enhancement to the sSFR, 0.15$-$0.2 dex. 
\item Further, in paired galaxies with large mass ratios, the less massive galaxies experience higher levels of sSFR enhancement than the more massive galaxy in the pair by $\sim$0.1 dex.
\item The sSFR enhancement increases with closer projected separations where galaxies within 5$-$15 kpc have their sSFR enhanced by 0.3$-$0.4 which is consistent with previous single fiber studies. 
\item When splitting the sSFR enhancement profiles by stellar mass, mass ratio, or projected separation, the general shape of the enhancement profile is unchanged. While lower projected separations or mass ratios closer to 1:1 may raise the level of the central enhancement, the profiles still linearly fall to zero enhancement around 1.0 R/\reff.
\item The paired galaxies will experience a fractional mass growth rate of 1$-$6\%, where the growth in mass is largest in low mass galaxies.
